\vfill World of Warcraft's New Way to Get You Hooked

### 2013-04-03 paper.tex
Cleaned up further, sadly OS X line endings have made everything change.

\section{Focus Question}
Is Blizzard's use of World of Warcraft: Starter Edition ethical? \\

World of Warcraft: Starter Edition could precipitate an ``outbreak'' of videogame addiction leading to terrible incidents. \cite{SnowlyDeath}\cite{ShawnDeath}\cite{ParentsBlameWoWForDeath}\cite{DeathAfter50Hours}\cite{YouStoleMySwordNowYouDie} Deaths and addiction linked to World of Warcraft would cause a sharp decline in the sale of games. Blizzard made \$4.85 billion last year and has around 4700 employees. \cite{BlizzardFinances}\cite{BlizzardEmployeeCount} At its worst the company could fail and those employees would be without income and looking for work.

@@

\subsection{How Long Does It Take To Hit The Cap?}
With World of Warcraft's Starter Edition, the most debilitating restriction is the level cap. Starter Edition accounts are capped at level 20. \cite{StarterEditionFAQ} With the current highest level at 90, the cap happens quite early in gameplay. I created two different characters and took them from level 1 to 20. It took me 6 and 8 hours respectively to reach the cap.\footnote{I am an expert player with over two months of in-game time logged on my account. A newer player would likely take longer than this to reach level 20.}
Some users believe they have become addicted to MMORPGs from the moment they first play them. \cite{JustAComputerGameRight} There appears to be no cut and dry length of time that it takes to become addicted to something, different people become addicted to things at different rates.
Because there's no evidence that time is a factor in becoming addicted, the real argument becomes whether or not most users become addicted at all, whether the dangers of becoming addicted are outweighed by other factors, and whether Blizzard is doing it's part to warn its users.

\subsection{SE Code Section 1.02\\Moderate the interests of the software engineer, the employer, the client and the users with the public good.}
In the next few sections this tenet is broken down into its parts. I will explore the ``interests of the software engineer,'' in this case World of Warcraft's software engineers, the ``interests of ... the employer,'' in this case Blizzard Entertainment, and lastly the ``interests of ... the client and the users,'' the player. Following that will be a discussion of the ``public good.''\\
In this situation SE Code Section 1.02 becomes: ``World of Warcraft's Engineers shall moderate the interests of themselves, Blizzard Entertainment, and the players of World of Warcraft with the public good.''
\subsubsection{Interests of the Software Engineer}
It is common sense that a software engineer strives to make working software. For World of Warcraft's software engineers, this success is measured by how many players enjoy new content and continue to play the game. Considerations are made for balance and gameplay as well as story. \cite{DevWatercooler} These engineers are also interested in keeping their jobs and continuing to make money. In order to keep their jobs, their interests must lie in getting players buy the software and then to come back to the game world often and play the game.
\subsubsection{Interests of Blizzard Entertainment}
Blizzard Entertainment is a publicly traded company. \cite{BlizzStock} As such its main goal is to show its shareholders that it is profiting and increasing the volume of business it handles. In no uncertain terms, Blizzard Entertainment wants as many people to buy and play their games, World of Warcraft included, as possible. Since World of Warcraft is subscription-based \cite{WoWSubscription}, Blizzard Entertainment earns money every single month from World of Warcraft players. None of the other games that Blizzard makes are subscription based. \cite{BlizzGameList} That means that Blizzard only ever receives one payment for any of its other games from customers. Blizzard Entertainment's interests are in getting more players to buy their games, particularly World of Warcraft which nets them money every month in addition to the initial cost.
\subsubsection{Interests of the Player}
By purchasing a game like World of Warcraft, players are committing themselves to pay Blizzard Entertainment every month. \cite{WoWSubscription} Some players are reaping benefits that they would not be able to get otherwise, such as an increase in social interactions. \cite{IsThereEvidenceOfInternetAddiction} Users who play the game rarely navigate the world alone, instead they form groups with other players, join guilds, and tackle quests and dungeons that require teamwork and talking to others.
One particular study found that MMORPG users play for a mixture of ten reasons: advancement, mechanics, competition, socializing, relationship, teamwork, discovery, role-playing, customization, and escapism. These ten reasons break down into three components: achievement, social, and immersion. \cite{PlayerMotivations} The achievement component encompasses achievement, ``the desire to gain power, progress rapidly, and accumulate in-game symbols of status or wealth,'' mechanics, ``having an interest in analyzing the underlying rules and system in order to optimize character performance,'' and competition, ``the desire to challenge and compete with others.'' The social component deals with socializing, ``having an interest in helping and chatting with other players,'' relationship, ``the desire to form long-term meaningful relationships with others,'' and teamwork, ``deriving satisfaction from being part of a group effort.'' Lastly, the immersion component is discovery, ``finding and knowing things that most other players don't know about,'' role-playing, ``creating a persona with a background story and interacting with other players to create improvised story,'' customization, ``having an interest in customizing the appearance of their character,'' and escapism, ``using the online environment to avoid thinking about real life problems.'' \cite{PlayerMotivations} (All information following in this section is from the same study. \cite{PlayerMotivations}) MMORPGs offer a sense of achievement through completing quests, challenging others, and optimizing character performance. MMORPGs offer the chance to socialize with people you've never met, join groups to accomplish tasks, and form relationships with people through guilds and group events. MMORPGs also offer a sense of immersion through their expansive worlds that can be explored, the opportunity to role-play with others, the ability to customize your characters in a nearly infinite set of ways, and the chance to step into a different world where your real-world problems fade away. The interests of the player boil down to achievements, social interaction, and immersion in a world outside their own.

>>>>

\section{Focus Question}
Is Blizzard's use of World of Warcraft: Starter Edition ethical?

@@

\subsection{SE Code Section 1.02\\Moderate the interests of the software engineer, the employer, the client and the users with the public good.}
In the next few sections this tenet is broken down into its parts. I will explore the ``interests of the software engineer,'' in this case World of Warcraft's software engineers, the ``interests of ... the employer,'' in this case Blizzard Entertainment, and lastly the ``interests of ... the client and the users,'' the player. Following that will be a discussion of the ``public good.''\\
In this situation SE Code Section 1.02 becomes: ``World of Warcraft's Engineers shall moderate the interests of themselves, Blizzard Entertainment, and the players of World of Warcraft with the public good.''

\subsubsection{Interests of the Software Engineer}
It is common sense that a software engineer strives to make working software. For World of Warcraft's software engineers, this success is measured by how many players enjoy new content and continue to play the game. Considerations are made for balance and gameplay as well as story. \cite{DevWatercooler} These engineers are also interested in keeping their jobs and continuing to make money. In order to keep their jobs, their interests must lie in getting players buy the software and then to come back to the game world often and play the game.

\subsubsection{Interests of Blizzard Entertainment}
Blizzard Entertainment is a publicly traded company. \cite{BlizzStock} As such its main goal is to show its shareholders that it is profiting and increasing the volume of business it handles. In no uncertain terms, Blizzard Entertainment wants as many people to buy and play their games, World of Warcraft included, as possible. Since World of Warcraft is subscription-based \cite{WoWSubscription}, Blizzard Entertainment earns money every single month from World of Warcraft players. None of the other games that Blizzard makes are subscription based. \cite{BlizzGameList} That means that Blizzard only ever receives one payment for any of its other games from customers. Blizzard Entertainment's interests are in getting more players to buy their games, particularly World of Warcraft which nets them money every month in addition to the initial cost.

\subsubsection{Interests of the Player}
By purchasing a game like World of Warcraft, players are committing themselves to pay Blizzard Entertainment every month. \cite{WoWSubscription} Some players are reaping benefits that they would not be able to get otherwise, such as an increase in social interactions. \cite{IsThereEvidenceOfInternetAddiction} Users who play the game rarely navigate the world alone, instead they form groups with other players, join guilds, and tackle quests and dungeons that require teamwork and talking to others.
One particular study found that MMORPG users play for a mixture of ten reasons: advancement, mechanics, competition, socializing, relationship, teamwork, discovery, role-playing, customization, and escapism. These ten reasons break down into three components: achievement, social, and immersion. \cite{PlayerMotivations} The achievement component encompasses achievement, ``the desire to gain power, progress rapidly, and accumulate in-game symbols of status or wealth,'' mechanics, ``having an interest in analyzing the underlying rules and system in order to optimize character performance,'' and competition, ``the desire to challenge and compete with others.'' The social component deals with socializing, ``having an interest in helping and chatting with other players,'' relationship, ``the desire to form long-term meaningful relationships with others,'' and teamwork, ``deriving satisfaction from being part of a group effort.'' Lastly, the immersion component is discovery, ``finding and knowing things that most other players don't know about,'' role-playing, ``creating a persona with a background story and interacting with other players to create improvised story,'' customization, ``having an interest in customizing the appearance of their character,'' and escapism, ``using the online environment to avoid thinking about real life problems.'' \cite{PlayerMotivations} (All information following in this section is from the same study. \cite{PlayerMotivations}) MMORPGs offer a sense of achievement through completing quests, challenging others, and optimizing character performance. MMORPGs offer the chance to socialize with people you've never met, join groups to accomplish tasks, and form relationships with people through guilds and group events. MMORPGs also offer a sense of immersion through their expansive worlds that can be explored, the opportunity to role-play with others, the ability to customize your characters in a nearly infinite set of ways, and the chance to step into a different world where your real-world problems fade away. The interests of the player boil down to achievements, social interaction, and immersion in a world outside their own.
